Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-07-18 Writing/NatureWater/1stRoundRevision/replies/reply_to_reviewer_1.rmd

---
output:
  pdf_document:
    keep_tex: true
---
Reviewer #1 (Comments for the Author):

This manuscript assesses the impact of aquifer thickness on agricultural production during drought in the area overlying the High Plains aquifer in the central United States. The authors use crop, meteorological, and aquifer thickness data in regressions to develop relationships among these quantities. Their most important finding is the impact of aquifer thickness on productivity during drought and the ramifications for aquifer conservation efforts. I have read this manuscript over a number of times and found it of considerable interest and worthy of publication in Nature Water. However, I do have a few concerns that I wanted to bring to the attention of the authors. My most important concerns are as follows (not necessarily in order of importance):

\textcolor{blue}{We really appreciate your comments. Addressing them has made our paper significantly stronger. Please find our replies to your comments written in blue.}

1. Aquifer thickness versus saturated thickness – The authors speak of "aquifer saturated thickness", which is a redundant term. Remember, the definition of an aquifer is a saturated unit – there is no unsaturated aquifer. Please use "aquifer thickness" to avoid propagating this confusing usage into the future.

\textcolor{blue}{All the instances of "saturated thickness" and "aquifer saturated thickness" have been replaced by "aquifer thickness".}

2. Dependence on aquifer thickness – I had a number of questions about the relationships displayed in Figure 1. First, the regression appears to have been done year by year. Are the quantiles recalculated each year to reflect that there may be changes between categories as time goes on? Second, when the authors state the lowest quantile goes from 0 to 33% are they assuming that the minimum thickness of 9 m is the zero point? Third, I assume that the county and year fixed effects are calculated as part of the regression. Are there any insights that can be gained from the interpretation of those quantities? Fourth, the authors need to point out that the same aquifer thickness can result in dramatically different transmissive characteristics. Thus, a relatively thin aquifer interval composed of coarse sands and gravels may be more resilient to drought than a much thicker aquifer dominated by fine sands and silts. Could that be a major player in the error bars in the plots? Please mention this at some point in the manuscript.

- First point: \textcolor{blue}{Quantiles are not recalculated each each. If we do so, then the same quantile would represent different aquifer thickness range every year. When modeling the impact of aquifer thickness, this is not desirable because we are implicitly making assumptions that different ranges of aquifer thickness would have the same impact.}
-  Second point: \textcolor{blue}{Actually, 0% quntile is 12 m. "9m" was repalced with "12m" in the current manuscript.}
-  Third point: \textcolor{blue}{No interesting insights will be gained from them because they are merely shifters without any intractions with aquifer thickness. They tell us how much higher yield in a single county relative to another county on average. But, we do not beleive that is useful information at least in this artcile.}
-  Fourth point: \textcolor{blue}{This is a very important point. We have added texts describing the implications of this fact in page xx.}

3. Conditioned on drought risk exposure – I wasn’t sure what the authors meant by this in the second paragraph of Section 2.2. This must be related to the water deficit but the lack of explanation in the main text or the methods section left me wondering. Please clarify at some point in the manuscript.

\textcolor{blue}{What we meant by "conditiond on drought risk exposure" is "after taking into account the impact of climate variables." In order to avoid confusion, we simply removed it from the sentence in the current manuscript.}

4. Equation 2 – Please define all terms in the equation and provide an explanation of what is meant by "estimated semi-parametrically".

\textcolor{blue}{We now have a more detailed mathematical expressions of the model with all the terms explained. We eliminated the word semi-parametrically as it is not ncecessary. Generalized additive model that estiamtes the impact of a variable using splines without functional form assumption are sometimes referred to semi-parametric regression. That's what we meant by "estimate semi-parametrically" in our previous manuscript.}

5. Kansas and California efforts – I’ve seen a number of papers describing efforts in Kansas and California to address aquifer depletion via LEMAs (Kansas) and SGMA (California). The 1985-2016 period examined by the authors was before most of the efforts in Kansas got started. My understanding is that the first LEMA was established in 2013 but then no more were set up until after 2016. This manuscript would be more impactful if the authors could speculate on how their relationships might change with greater attention to groundwater conservation. For example, I would think the decrease in yield the authors observe with negative water deficits would disappear when the irrigators are using soil moisture sensors and are focused on water-use efficiency. I would expect changes as well for tipping points.

\textcolor{blue}{Tim?}

## Minor points

+ Section 1, 3rd Paragraph – "Two aquifer pathways exist through aquifer depletion will.." – confusing sentence – please replace "through" with "by which".

\textcolor{blue}{Corrected.}

+ Section 1, 4th Paragraph – "we generate empirical evidence" seems awkward to me. How about replacing with "we present empirical evidence"? Next sentence – "Our analysis advances on previous…" reads better as "Our analysis extends previous…" There are many issues such as these and the previous comment. Please address such issues so that the manuscript is an easier read.

\textcolor{blue}{We completely agree. Both of the suggested changes have been made.}

+ Section 2.1 title – I would delete "marginally" as it is not needed and some will find it confusing.

\textcolor{blue}{Tim, any thoughts on this. I feel like it is still important to say the increase is not much.}

+ Figure 2 - Why do the plots stop at a thickness of about 12 m when you had earlier defined the minimum thickness as 9 m and mention 10 m in the text when discussing the relationships?

\textcolor{blue}{Thank you for catching this. 9m and 10m should have been 12m.}

+ Section 2.3, First Paragraph – please clarify that "average per-area production" is a combination of irrigated and rainfed yields. That would make the manuscript easier to follow.

\textcolor{blue}{We added "(weighted-average of irrigated and rainfed yields, where weights are production area shares)" so that it is clear what average per-area production means.}

+ Figure 5 – I assume that the left plot is corn and the right soybean, but please clarify.

\textcolor{blue}{Thanks for noticing this. Fixed.}

+ Data and Code Availability – Are the data from Dr. Haacker on the Github repository? Please place them there prior to publication so future investigators can access them rather than having to track down Dr. Haacker. The data set is over seven years old so it should be made available according to the principles of open science.

\textcolor{blue}{Yes, we will make the data available online.}

+ Line before Equation (2) – "REF" should be replaced with the actual reference.

\textcolor{blue}{Tim?}

+ Discussion, First Paragraph – "reduction in aquifer depletion" should be "reduction in aquifer thickness".

\textcolor{blue}{Thank you for catching this. Fixed.}

+ Section 4.2, 3rd Paragraph – "the the impact" – remember to spell and grammar check prior to submission.

\textcolor{blue}{Thank you for catching this. Fixed.}

+ Reference in Section 3 – I doubt that "Butler Jr" is the surname, please replace with "Butler". Suffixes apparently aren’t handled well by some of the bibliography generation programs.

\textcolor{blue}{}

+ Please remember to close all parentheses and don't mispair brackets and parentheses. Such stuff can raise questions about the care used in the analysis.

### 2023-07-18 Writing/NatureWater/1stRoundRevision/replies/reply_to_reviewer_1.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

Reviewer \#1 (Comments for the Author):

This manuscript assesses the impact of aquifer thickness on agricultural
production during drought in the area overlying the High Plains aquifer
in the central United States. The authors use crop, meteorological, and
aquifer thickness data in regressions to develop relationships among
these quantities. Their most important finding is the impact of aquifer
thickness on productivity during drought and the ramifications for
aquifer conservation efforts. I have read this manuscript over a number
of times and found it of considerable interest and worthy of publication
in Nature Water. However, I do have a few concerns that I wanted to
bring to the attention of the authors. My most important concerns are as
follows (not necessarily in order of importance):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Aquifer thickness versus saturated thickness -- The authors speak of
  ``aquifer saturated thickness'', which is a redundant term. Remember,
  the definition of an aquifer is a saturated unit -- there is no
  unsaturated aquifer. Please use ``aquifer thickness'' to avoid
  propagating this confusing usage into the future.
\end{enumerate}

\textcolor{blue}{ferkjnf}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\item
  Dependence on aquifer thickness -- I had a number of questions about
  the relationships displayed in Figure 1. First, the regression appears
  to have been done year by year. Are the quantiles recalculated each
  year to reflect that there may be changes between categories as time
  goes on? Second, when the authors state the lowest quantile goes from
  0 to 33\% are they assuming that the minimum thickness of 9 m is the
  zero point? Third, I assume that the county and year fixed effects are
  calculated as part of the regression. Are there any insights that can
  be gained from the interpretation of those quantities? Fourth, the
  authors need to point out that the same aquifer thickness can result
  in dramatically different transmissive characteristics. Thus, a
  relatively thin aquifer interval composed of coarse sands and gravels
  may be more resilient to drought than a much thicker aquifer dominated
  by fine sands and silts. Could that be a major player in the error
  bars in the plots? Please mention this at some point in the
  manuscript.
\item
  Conditioned on drought risk exposure -- I wasn't sure what the authors
  meant by this in the second paragraph of Section 2.2. This must be
  related to the water deficit but the lack of explanation in the main
  text or the methods section left me wondering. Please clarify at some
  point in the manuscript.
\item
  Equation 2 -- Please define all terms in the equation and provide an
  explanation of what is meant by ``estimated semi-parametrically''.
\item
  Kansas and California efforts -- I've seen a number of papers
  describing efforts in Kansas and California to address aquifer
  depletion via LEMAs (Kansas) and SGMA (California). The 1985-2016
  period examined by the authors was before most of the efforts in
  Kansas got started. My understanding is that the first LEMA was
  established in 2013 but then no more were set up until after 2016.
  This manuscript would be more impactful if the authors could speculate
  on how their relationships might change with greater attention to
  groundwater conservation. For example, I would think the decrease in
  yield the authors observe with negative water deficits would disappear
  when the irrigators are using soil moisture sensors and are focused on
  water-use efficiency. I would expect changes as well for tipping
  points.
\end{enumerate}

\hypertarget{minor-points}{%
\subsection{Minor points}\label{minor-points}}

Section 1, 3rd Paragraph -- ``Two aquifer pathways exist through aquifer
depletion will..'' -- confusing sentence -- please replace ``through''
with ``by which''.

Section 1, 4th Paragraph -- ``we generate empirical evidence'' seems
awkward to me. How about replacing with ``we present empirical
evidence''? Next sentence -- ``Our analysis advances on
previous\ldots{}'' reads better as ``Our analysis extends
previous\ldots{}'' There are many issues such as these and the previous
comment. Please address such issues so that the manuscript is an easier
read.

Section 2.1 title -- I would delete ``marginally'' as it is not needed
and some will find it confusing.

Figure 2 - Why do the plots stop at a thickness of about 12 m when you
had earlier defined the minimum thickness as 9 m and mention 10 m in the
text when discussing the relationships?

Section 2.3, First Paragraph -- please clarify that ``average per-area
production'' is a combination of irrigated and rainfed yields. That
would make the manuscript easier to follow.

Figure 5 -- I assume that the left plot is corn and the right soybean,
but please clarify.

Data and Code Availability -- Are the data from Dr.~Haacker on the
Github repository? Please place them there prior to publication so
future investigators can access them rather than having to track down
Dr.~Haacker. The data set is over seven years old so it should be made
available according to the principles of open science.

Line before Equation (2) -- ``REF'' should be replaced with the actual
reference.

Discussion, First Paragraph -- ``reduction in aquifer depletion'' should
be ``reduction in aquifer thickness''.

Section 4.2, 3rd Paragraph -- ``the the impact'' -- remember to spell
and grammar check prior to submission.

Reference in Section 3 -- I doubt that ``Butler Jr'' is the surname,
please replace with ``Butler''. Suffixes apparently aren't handled well
by some of the bibliography generation programs.

Please remember to close all parentheses and don't mispair brackets and
parentheses. Such stuff can raise questions about the care used in the
analysis.

\end{document}

### 2023-07-18 Writing/NatureWater/1stRoundRevision/revised_manuscript/cover_letter.tex
revision started

\title{The Impacts of Aquifer Depletion on the Ability of Groundwater Irrigation to Buffer Against Undesirable Weather}
\author{Taro Mieno\footnote{Department of Agricultural Economics, University of Nebraska Lincoln, \href{mailto:tmieno2@unl.edu}{\nolinkurl{tmieno2@unl.edu}}} \and Timothy Foster\footnote{Department of WWW, University of Manchester, \href{mailto:timothy.foster@manchester.ac.uk}{\nolinkurl{timothy.foster@manchester.ac.uk}}} \and Shunkei Kakimoto\footnote{Department of Applied Economics, University of Minnesota} \and Nicholas Brozovic\footnote{Department of Agricultural Economics, University of Nebraska Lincoln, \href{mailto:nbrozovic@nebraska.edu}{\nolinkurl{nbrozovic@nebraska.edu}}}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Abstract goes here. It should be a single paragraph of around 200 words.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Groundwater-irrigated agriculture supports X\% of crop production worldwide and Y\% in U.S. (citations) It is a vital natural resource that brings tremendous economic and social benefits to the society. However, groundwater has been depleting rapidly all over the world due to intensive groundwater extraction that far exceeds the natural replenishment rate (citation). The importance of irrigation has been consistently growing over the years (Irmak).

\begin{itemize}
\tightlist
\item
  some mentionin of food security issues
\end{itemize}

The vast majority of studies on the climate change impacts on agricultural productivity focus on dryland production as it is far vulnerable to climate variations than irrigated production is. Another strand of literature seeks to understand the impact of climate change on water demand. Only a small collection of studies estimated the climate change impacts on irrigated production. It first seems that we do not need to understand the impact of climate change on irrigated production because irrigation provide a protection against climate variation. This would be true under the condition that farmer can apply desired amount of water without any physical or economic constraints, in which case, farmers can always take advantage of the benefit of irrigation to its full potential. Unfortunately, that is not the case for groundwater irrigated agriculture. Specifically, the stock of groundwater in the aquifer, measured as saturated thickness, affects the physical and economic constraints under which farmers operate. A decline in saturated thickness place stronger constraints on farmers and hurt the profitability of irrigated production.

There are mainly two reasons that saturated thickness affects the ability of aquifer to protect farmers from undesirable production conditions. First, as saturated thickness goes down due to groundwater depletion, the economic cost of pumping water increases. This affects overall agricultural productivity through two channels: extensive and intensive margins. When the cost of groundwater extraction becomes too high, irrigated production is no longer more profitable than dryland production. In this case, farmers will switch to irrigated production to dryland production even though groundwater still exists in the aquifer, which leads to significant reduction in crop yield (extensive margin). For those who have decided to irrigate, they would use less water as the cost of groundwater pumping is higher, thus lowering crop yield (intensive margin). Second and more important, well yield (a measure of how much water can be pumped per unit of time) declines as saturated thickness goes down (). This physical constraint has been shown to have far greater impacts on irrigation decisions than pumping cost \citep{mieno2021}. Similarly to the rising pumping cost case, declining well yield affects agricultural productivity through two channels. Too low a well yield makes producers shift to dryland production. For those irrigating, a lower well yield would lead to lower crop yield.

\citet{sampson2019} estimated the impact of saturated thickness on land value.

The vast majority of existing body of studies focus on estimating the impact of climate on dryland production. A small amount of studies that look at the impact of climate on irrigated production does not take into account saturated thickness. Our study fills this important gap and sheds light on the value of groundwater by recognizing the importance of saturated thickness and quantifying its ability to protect agricultural production from undesirable weather outcomes.

\citep{rad2020effects}
\citep{foster2014modeling}
\citep{foster2015analysis}

\hypertarget{materials-and-method}{%
\section{Materials and Method}\label{materials-and-method}}

In order to understand the impact of saturated thickness on agricultural production, this study examines the two channels through which saturated thickness affects average corn and soybean yields using econometric analysis. First, we quantify the impact of water deficit (\(WB\)) during the production season (May through September) on irrigated and dryland yields (\(Y\)) differentiated by the level of saturated thickness (\(ST\)). Mathematically, we estimate the expected value of yield conditional on water deficit, saturated thickness, and whether irrigated or not (\(I\)).

\[
E[Y_{i,t}|WB_{i,t}, ST_{i,t}, I_{i,t}]= f(WB_{i,t}, ST_{i,t}, I_{i,t})
\]

where \(i\) indicates county and \(t\) indicates year.

\noindent Second, we estimate the expected value of the share of irrigated acres as a function of saturated thickness and average water deficit (averaged over years). Historical average water deficit is used in place of concurrent water deficit \(WB_{i,t}\) because the decision of whether to irrigate or not is made based on what the history of crop water demand rather than the concurrent weather that is yet to unfold at the time of decision making.

\[
E[S_{i,t}|WB_{i,t}, ST_{i,t}]= g(\overline{WB}_i, ST_{i,t}) 
\]

\noindent Once they are estimated, we can estimate the average crop yield as the share-weighted average of dryland and irrigated yields as follows:

\[
\hat{h}(WB_t, St_t)= (1 - \hat{g}(\overline{WB}, ST) )\cdot \hat{f}(WB_t, ST_t, I_t = 0)+ \hat{g}(\overline{WB}, ST) \cdot \hat{f}(WB_t, ST_t, I_t = 1)
\]

\noindent where \(\hat{f}(\cdot)\) and \(\hat{g}(\cdot)\) are estimated version of \(f(\cdot)\) and \(g(\cdot)\), respectively. For example, the impact of a change in saturate thickness level, say from \(ST_{before}\) to \(ST_{after}\), on crop yield is then \(\hat{h}(WB_t, St_{before})\) - \(\hat{h}(WB_t, St_{before})\).

\hypertarget{econometric-models}{%
\subsection{Econometric Models}\label{econometric-models}}

We use the same model specification for corn and soybean yield regression analysis. The estimating equation is written as follows:

\[
y_{i,t} = \sum_{j=1}^4 \sigma_j C_j + \sum_{j=1}^4 \sum_{k}^K \beta_{j,k}\phi_{j,k}(WB_{i,t})\cdot C_j  + \alpha_i + \phi_t + v_{i,t}
\]

Individual county and year are indicated by \(i\) and \(t\), respectively. In the model, \(y_{i,t}\) is crop yield for county \(i\) in year\(t\). \(C_j\) is a dummy variable that takes 1 if observation \(i\) is under category \(C_j\), 0 otherwise. The observations are grouped into four categories: dryland production, observations with saturated thickness falling under the \([0-33.3\%)\)j, \([33.3-66.7\%)\), and \([66.7\% - 100\%]\) quantile ranges of saturated thickness. The three irrigated production groups have about the same number of observations, which ensures statistical stability in estimating yield response to drought by category. \(\sigma_j\) represents the category specific intercept. \(WB_{i,t}\) is water deficit defined as total reference evapotranspiration less precipitation between May through September. Category-specific impact of water deficit on yield is represented by \(\sum_{k}^K \beta_k\phi_{j,k}(WB_{i,t})\), where \(\phi_{j,k}\) is \(k\)th spline basis function and \(\beta_k\) is its coefficient. Spline functions are used to capture the potential non-linearity of the impact of water deficit on yield in a flexible manner. The model can be interpreted as hierarchical generalized additive model, where each of saturated thickness level category has its own yield response function to drought (cite). To control for time-invariant heterogeneity across counties and yearly shocks, county fixed effects (\(\alpha_i\)) and year fixed effects (\(\phi_t\)) are included. Standard error estimation is clustered by county.

The impact of saturated thickness on the share of irrigated acres is estimated using a fractional logit model under the generalize additive model framework in which the impact of saturated thickness, average water deficit, and their interactions are estimated semi-parametrically, allowing for flexible representation of their impacts. The estimating equation is as follows:

\[
log(\frac{s_{i,t}}{1-s_{i,t}}) = \beta_0 + h(ST_{i,t}, \overline{WB}_i) + \mu_t + v_{i,t}
\]

, where \(s_{i,t}\) is the share of irrigated production. \(h(ST_{i,t}, \overline{WB}_i)\) consists of basis functions and their coefficients to repsent the the impact of \(ST_{i,t}\), \(\overline{WB}_i\), and their interactions (see Appendix ?? for further details of how \(h(ST_{i,t}, \overline{WB}_i)\) are mathematically defined).

\hypertarget{data}{%
\subsection{Data}\label{data}}

Figure \ref{fig:geo-focus} shows the geographical coverage of the data for regression analysis for corn and soybean.

\begin{figure}

{\centering \includegraphics[width=432px]{manuscript_DR_files/figure-latex/geo-focus-1}

}

\caption{Geographic focus of the study for corn and soybean}\label{fig:geo-focus}
\end{figure}

First, county-level irrigated and dryland yield data for corn and soybean for 1985 through 2016 were obtained from the USDA NASS QuickStat. Both dryland and irrigated yield observations are used for counties that shares more than 75\% of their area with the Ogallala aquifer. Only dryland yield observations are used for counties that do not satisfy the condition. We use dryland yield data outside of the Ogallala aquifer to supplement dryland yield data for more accurate estimation of the impact of water deficit for dryland production. Daily precipitation and reference ET were obtained from gridMET \citep{Abatzoglou2013}. Water deficit (\(WB_{i,t}\)) was then calculated by subtracting total precipitation from total reference evapotranspiration from May through September. Saturated thickness data were obtained from Dr.~Erin Haacker \citep{haacker2016water}. Observations that have saturated thickness value lower than 12 meter were eliminated as saturated thickness below that is considered not suitable for irrigation (Tim citations here).

Figure \ref{fig:deficit-hist} presents annual water deficit by year for the corn and soybean data. Water deficit is highly variable year to year. Year 2012 in particular stands out as an extremely dry year in the region. We use the year-to-year variations in water deficit to identify its impact on crop yield. Figure \ref{fig:yield-hist} compares historical corn and soybean yields for dryland and irrigated production. The power of irrigation is clearly evident in this figure. First, irrigation provides protection against severe drought years like 2000, 2002, 2012. While the dryland yield dives down in those years, irrigated yields is almost unaffected by the severe droughts. First key objective of our study is to see if and how much the ability of irrigation is diminished when saturated thickness goes down. Figure \ref{fig:sat-dist} presents the distribution of saturated thickness. Saturated thickness vary substantially, but most of the sites have saturated thickness of less than 50 meters. We exploit the observed variations in saturated thickness to identify its influence on the ability of irrigation to drought. Second, irrigated yield is always about 50\% or higher than dryalnd yield irrespective of the degree of water deficit. We will investigate the negative impact of a decline in saturated thickness on the percentage of irrigated fields, thus reducing crop yields.

\begin{figure}

{\centering \includegraphics[width=432px]{manuscript_DR_files/figure-latex/deficit-hist-1}

}

\caption{Distribution of water deficit for corn and soybean datasets}\label{fig:deficit-hist}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=432px]{manuscript_DR_files/figure-latex/yield-hist-1}

}

\caption{Historical dryland and irrigated yields for corn and soybean}\label{fig:yield-hist}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=432px]{manuscript_DR_files/figure-latex/sat-dist-1}

}

\caption{Distribution of saturated thickness}\label{fig:sat-dist}
\end{figure}

\hypertarget{results-and-discussions}{%
\section{Results and Discussions}\label{results-and-discussions}}

\hypertarget{the-impact-of-drought-on-dryland-and-irrigated-yield}{%
\subsection{The impact of drought on dryland and irrigated yield}\label{the-impact-of-drought-on-dryland-and-irrigated-yield}}

Regression results are presented graphically rather than in tables because individual coefficients estimates on spline basis functions cannot be readily interpreted and also because visualized results are much easier to interpret. Figure \ref{fig:yield-response} shows yield response to water deficit for dryland production and groundwater-irrigated production at different levels of saturated thickness categories for corn (1st row) and soybean (2nd row).

\begin{figure}

{\centering \includegraphics[width=432px]{manuscript_DR_files/figure-latex/yield-response-1}

}

\caption{The impact of water deficit on dryland and irrigated yield by crop}\label{fig:yield-response}
\end{figure}

In a fairly wet year without much water stress, dryland and irrigated production result in very similar crop yields for both soybean and corn yields. Irrigated yields are only slightly higher than dryland yields at a negative value of water deficit (precipitation is greater than reference evapotranspiration). However, dryland yield declines sharply and steadily as water deficit increases. On the contrary, irrigated corn yield actually increase as water deficit increases up to about water deficit of 400mm at all the saturated thickness levels for both corn and soybean. This is because irrigation can turn unfavorable high-heat conditions into favorable growing conditions up to a point (Tim, citations?). However, yield response to water deficit departs after deficit of about 400mm depending on the level of saturated thickness. While higher saturated thickness categories can withstand water deficit and crop yield stays about the same for higher water deficit values, the low saturated thickness category experience non-trivial declines in crop yields as water deficit increases.

While there are some gaps in crop yield between the lowest and the higher saturated thickness categories, the lowest category still performs quite well for both corn and soybean. This is because the samples in irrigation in the regression analysis are those who opted to stay irrigated even when the saturated thickness level is low because irrigation is still powerful enough to provide protections for farmers. So, fields who would have experienced significant declines in crop yield in severe droughts due to low saturated thickness have opted out of irrigation and thus not observable in the data for irrigated production. Therefore, In order to capture the fuller impact of a decline in saturated thickness, it is important to understand how saturated thickness affects the decision of whether to irrigate or not.

\hypertarget{the-impact-of-saturated-thickness-on-the-share-of-irrigated-production}{%
\subsection{The impact of saturated thickness on the share of irrigated production}\label{the-impact-of-saturated-thickness-on-the-share-of-irrigated-production}}

\begin{figure}

{\centering \includegraphics[width=432px]{manuscript_DR_files/figure-latex/ir-share-1}

}

\caption{The impact of saturated thickness on the share of irriagate production by crop }\label{fig:ir-share}
\end{figure}

As shown in \ref{fig:ir-share}, saturated thickness has significant impacts on the share of irrigated production for corn. As saturated thickness declines, the share of irrigated acres for corn decreases. As saturated thickness declines, more land will be under dryland production, which is more vulnerable to drought, damaging average corn yield. Unlike corn, a decline in saturated thickness does not have as large negative impact on the share of irrigated production for soybean as corn. This is likely because soybean is less susceptible to drought compared to corn.

\hypertarget{total-impact-of-decline-in-saturated-thickness}{%
\subsection{Total impact of decline in saturated thickness}\label{total-impact-of-decline-in-saturated-thickness}}

Taking into account both impacts of saturated thickness, figure \ref{fig:tot-impact} shows how water deficit affects yield at different saturated thickness levels evaluated. For corn, in years with water deficit of 400mm or less saturated thickness does not have economically significant negative impacts because the difference in irrigated and dryland yields are small and also because irrigated yields are similar irrespective of the level of saturated thickness. The same can be said for soybean except that the soybean is more tolerant than corn and the saturated thickness starts to make differences at water deficit of 500mm of higher. In the year with water deficit of 800mm, average corn yield at saturated thickness of 186 meters is about 15\% and 25\% higher than at 75.7 meters and 29.6 meters, respectively. These findings illustrate that vulnerability to drought is heterogeneous among irrigated production depending on the level of saturated thickness. In summary, our studies confirms that a decline in saturated thickness makes yield production much more vulnerable to severe drought, and its negative impact is non-negligible.

It is also noteworthy that the impact of saturated thickness on crop yield in drought years is non-linear. Specifically, according to our estimates, the decline of saturated thickness from 186 meters to 75 meters would cause about one tonne/ha of damage in corn yield at water deficit of 1100mm. However, the decline of saturated thickness from 75 meters to 30 meters would cause about 1.5 tonne/ha of damage in corn yield. This means that the negative impact of a meter of decline in saturated thickness becomes larger as the aquifer depletion progresses. This has an important policy implication. Suppose we intend to limit groundwater pumping to maintain the stable level of groundwater stock by matching the natural replenishment rate for the purpose of sustainable groundwater-irrigated production. Then, we need to act early before saturated thickness becomes too low so that we do not need to product at a less productive steady state. The longer you wait, the harsher damage is going to be when you finally decide to place a policy that limits groundwater pumping.

\begin{figure}

{\centering \includegraphics[width=432px]{manuscript_DR_files/figure-latex/tot-impact-1}

}

\caption{The impact of water deficit on average crop yield by sturated thickness}\label{fig:tot-impact}
\end{figure}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This article quantified the impact of saturated thickness on average crop yield\ldots{}

\clearpage

\hypertarget{appendix-appendix}{%
\appendix}

\hypertarget{use-spei-in-place-of-water-deficit}{%
\section{Use SPEI in place of water deficit}\label{use-spei-in-place-of-water-deficit}}

Some studies use SPEI to represent drought severity. We also considered using SPEI, but the regressions with water deficit provides a better fit for both corn and soybean (see \ref{tab:fit-reg}).

\begin{table}[!h]

\caption{\label{tab:fit-reg}Comparison of R-squared of the regressions with water deficit and SPEI for corn and soybean}
\centering
\begin{tabular}[t]{lrr}
\toprule
Crop & SPEI & Water Deficit\\
\midrule
Corn & 0.899 & 0.903\\
Soybean & 0.840 & 0.846\\
\bottomrule
\end{tabular}
\end{table}

Qualitatively, the results with SPEI are similar to the regression results with water deficit as can be seen in \ref{fig:spei}.

\begin{figure}

{\centering \includegraphics[width=360px]{manuscript_DR_files/figure-latex/spei-1}

}

\caption{The impact of SPEI on crop yield for corn and soybean}\label{fig:spei}
\end{figure}

\clearpage

  \bibliography{DRA.bib}

>>>>

\title{Aquifer depletion exacerbates agricultural drought losses in the US High Plains}
\author{Taro Mieno\footnote{Department of Agricultural Economics, University of Nebraska Lincoln, \href{mailto:tmieno2@unl.edu}{\nolinkurl{tmieno2@unl.edu}}} \and Timothy Foster\footnote{Department of Mechanical, Aerospace and Civil Engineering, University of Manchester, \href{mailto:timothy.foster@manchester.ac.uk}{\nolinkurl{timothy.foster@manchester.ac.uk}}} \and Shunkei Kakimoto\footnote{Department of Applied Economics, University of Minnesota, \href{mailto:kakim002@umn.edu}{\nolinkurl{kakim002@umn.edu}}} \and Nicholas Brozovic\footnote{Department of Agricultural Economics, University of Nebraska Lincoln, \href{mailto:nbrozovic@nebraska.edu}{\nolinkurl{nbrozovic@nebraska.edu}}}}
\date{}

\begin{document}
\maketitle

Dear Editor,\\

Thank you for considering this manuscript for publication at Nature Water. In this study, we demonstrate the importance of aquifer conservation for ensuring agricultural resilience to current and future drought events. Using the High Plains Aquifer in the central United States as a case study, we show that declines in aquifer saturated thickness significantly reduces the overall productivity of both corn and soybean during drought events. Negative impacts on agricultural productivity during droughts occur both through reductions in per-area crop yields, and reductions in expected irrigated areas as farmers are forced to transition land to rainfed production as a result of water supply constraints caused by aquifer depletion. Our findings demonstrate that unsustainable groundwater use is likely to substantially increase vulnerability of agriculture to drought in many parts of the world, and highlight the need for greater focus on proactive conservation of groundwater stocks to support adaptation of farms, agricultural supply chains, and rural economies to climate change.\\

Our work is significant because we quantify the impact of a decline in saturated thickness of groundwater on the resiliency of agricultural production, which has rarely been quantified in the literature before. We confirm quantitatively the importance of keeping relatively high saturated thickness for sustainable high-productivity agricultural production. While our results are for the US High Plains, the insights we generated here are likely to hold elsewhere.\\

This article does not have any realated manuscripts that are under consideration. However, its pre-print has been published at EarthArXiv (\url{https://eartharxiv.org/repository/view/5126/}). We have not had any prior discussions with a Nature Water editor about this work. \\

Here is the list of potential reviewers that we may recommend.

\begin{itemize}
    \item James Butler (\url{https://geo.ku.edu/people/james-k-butler-jr})
    \item Marc Muller (\url{https://engineering.nd.edu/faculty/marc-francois-muller/})
    \item Sam Zipper (\url{https://www.samzipper.com/authors/samuel-c.-zipper/})
    \item Christian Klassert (\url{https://www.ufz.de/index.php?en=38767})
    \item Jen Burney (\url{http://www.jaburney.net/})
\end{itemize}

We do not have any reviewers that we would like to avoid.
